### 2018-01-02 ident/docs/ident_manuscript.tex

\begin{document}
\section{Introduction:}	
The dynamic response of any system can be modeled as a set of ordinary differential equations (ode) (Equation \ref{eq:output}a) in the state variables ($\mathbf{x}$). The changes in the state variables are expressed as a function of the state variables, the system parameters ($\theta$) and the inputs to the system ($u$).	
\begin{subequations}\label{eq:output}
	\begin{align}
	\dot{\mathbf{x}} = g(\mathbf{x}, \theta, u)\\
	\mathbf{y} = h(\mathbf{x},\mu,u)
	\end{align}
\end{subequations}
When not all states are observable, Equation (\ref{eq:output}a) is usually augmented with a system that defines the relationship between the observable output variables ($\mathbf{y}$) and the state variables ($\mathbf{x}$) as in Equation (\ref{eq:output}b). The parameters used to establish this relationship (Equation \ref{eq:output}b), $\mu$, may or may not include system parameters $\theta$ defined in Equation (\ref{eq:output}a).

@@

\section{Methods:}
\subsection{Kinetic model of gluconeogenesis in E. coli:}
The proposed model for acetate consumption through gluconeogenesis and its corresponding kinetic model is used as a case study to illustrate the utility of identifiability analysis for the design of experiments for estimating parameters in kinetic models of metabolism. The kinetic model is described below.
\begin{equation}\label{eq:ode1}
\frac{d}{dt}pep=v_1-v_2-v_4
\end{equation}
\begin{equation}\label{eq:ode2}
\frac{d}{dt}fdp=v_2-v_3
\end{equation}
\begin{equation}\label{eq:ode3}
\frac{d}{dt}E=v_{e,max}\left(\frac{1}{1+\left(\frac{fdp}{K_{e}^{fdp}}\right)^{n_e}}\right) - d E
\end{equation}
The kinetic expressions for fluxes $v_1$ through $v_4$ are given below. The consumption of acetate through $v_1$ and conversion of \textit{pep} through $v_2$ are expressed in Equations (\ref{eq:flux1}) and (\ref{eq:flux2}) respectively using Michaelis-Menten kinetics. The acetate flux through $v_1$ is also governed by the quantity of available enzyme E. 
\begin{equation}\label{eq:flux1}
v_1 = k_{1}^{cat}E\frac{acetate}{acetate+K_{1}^{acetate}}
\end{equation}	
\begin{equation}\label{eq:flux2}
v_2 = V_{2}^{max}\frac{pep}{pep+K_{2}^{pep}}
\end{equation}
\begin{equation}\label{eq:flux3}
v_3 = V_{3}^{max}\frac{\tilde{fdp}\left(1+\tilde{fdp}\right)^3}{\left(1+\tilde{fdp}\right)^4+L_3\left(1+\frac{pep}{K_{3}^{pep}}\right)^{-4}}
\end{equation}
The allosterically regulated flux $v_3$ for the consumption of \textit{fdp} is expressed in Equation (\ref{eq:flux3}) using the Monod-Wyman-Changeux (MWC) model for allosterically regulated enzymes, where $\tilde{fdp}$ refers to the ratio of \textit{fdp} with respect to its allosteric binding constant $K_{3}^{fdp}$. The added flux $v_4$ for the export of \textit{pep} is expressed as a linear equation dependent on $pep$ in Equation (\ref{eq:flux4}).
\begin{equation}\label{eq:flux4}
v_4 = k_{4}^{cat}.pep
\end{equation}

\subsection{A method for practical identifiability of kinetic models of metabolism:}
In this section, we show how practical identifiability of kinetic parameters in a dynamic model of metabolism can be established using the algebraic relationship between fluxes that form the network.
In a kinetic model, the value of every flux $v_i$ is expressed using one of the many available enzyme kinetic formulations. Without loss of generality, all of these kinetic formulations can be expressed as nonlinear algebraic equations (Equation \ref{eq:flux_gen}). The fluxes are expressed as a function of the metabolite concentrations $x$ and the kinetic parameters $\theta$. 
\begin{align}\label{eq:flux_gen}
v = f(\mathbf{x},\mathbf{\theta})=\frac{N(\mathbf{x},\theta)}{D(\mathbf{x},\theta)}
\end{align}
In Equation \ref{eq:flux_gen}, $N(\mathbf{x},\theta)$ is the numerator of $f$, and $D(\mathbf{x},\theta)$ is the denominator of $f$.

Let $\theta \in \mathbb{R}^p$ for each flux $v_i$ in the network. Also, let $E\in \mathbb{R}^n$ (and $j = {1, 2, ..., n}$) be the set of all possible experiments from which measurements of metabolite concentrations $\mathbf{x}_j$ and reaction fluxes $\mathbf{v}_j$ are available. For a given flux $v_i$, if all parameters $\theta \in \mathbb{R}^p$ are identifiable, then

if all parameters are identifiable, theoretically, the $P$ unknown parameters in $\theta$ can be estimated by solving $P\subset{1, 2, ..., n}$ nonlinear algebraic equations shown in Equation \ref{eq:nonlineq}.
\begin{align}\label{eq:nonlineq}
v_k = \frac{N(\mathbf{x}_k,\theta)}{D(\mathbf{x}_k,\theta)} && \forall k\in{1, 2, ..., k, ..., P}
\end{align}
Here, $K = {1, 2, ..., k, ..., P}$ is the index set of $P$ experiments for which metabolite concentration ($\mathbf{x}_K$) and flux ($\mathbf{v}_K$) measurements are available for each metabolite and flux within the metabolic network. Parameter $\theta_p\in\theta$ is said to be identifiable if

However, if for any index $l$, $\mathbf{x}_l$ and $\mathbf{v}_l$ do not satisfy the condition that $D(\mathbf{x}_l,\theta) \neq 0$, i.e., at least one flux $v_i$ is not identifiable based on experiment $l$, then the $l$th experiment cannot be used for parameter estimation and should be discarded. Consequently, an additional experiment $m \in {1, 2, ..., n}, m\neq l$ is needed to estimate all $p$ parameters in $\theta$. This analysis can be performed for each flux in a metabolic network independent of all the other fluxes. This enables this method to be scalable to even genome-scale models. The following section demonstrates this methodology for one of the fluxes in the gluconeogenic model described earlier.

\subsection{Experimental design for parameter estimation in kinetic models of metabolism:}
The identifiability of each parameter based on each experiment indexed as $j = {1, ..., n}$

\subsection{Identifiability of parameters in a kinetic model of gluconeogenesis:}

@@

@@

We use flux $v_2$ to demonstrate the identifiability analysis method described in the previous section. Flux $v_2$ has two parameters, $V_2^{max}$ and $K_2^{pep}$ that need to be estimated from experimental data. Here, we assume that at least two different sets of experimental data for the concentrations and fluxes are available. Accordingly, we label these dataset as ${pep}^1$, ${v_2}^1$ and ${pep}^2$, ${v_2}^2$ respectively. Subsequently, these experimental datasets can be included in the model to form two simultaneous nonlinear algebraic equations in the parameters $V_2^{max}$ and $K_2^{pep}$ (Equation \ref{eq:v2_par}).
\begin{subequations}\label{eq:v2_par}
	\begin{align}		
	V_2^{max} = \frac{v_2^1v_2^2(pep^1-pep^2)}{v_2^2pep^1-v_2^1pep^2}\\
	K_2^{pep} = \frac{pep^1(v_2^1pep^2-v_2^2pep^2)}{v_2^2pep^1-v_2^1pep^2}
	\end{align}
\end{subequations}

@@

\section{Results:}
\subsection{Getting closed-form expressions for each flux in Kotte model:}
\subsection{Experiments that can establish identifiability for each flux in Kotte model:}	
\subsection{Combinations of experiments that will enable identification of all model parameters:}
\subsection{Expanding methodology to RBC model:}

>>>>

\begin{document}
\section{Introduction:}	
The dynamic response of any system can be modeled as a set of ordinary differential equations (ode) (Equation \ref{eq:output}a) in the state variables ($\mathbf{x}$). The changes in the state variables are expressed as a function of the state variables, the system parameters ($\theta$) and the inputs to the system ($u$).	
\begin{subequations}\label{eq:output}
	\begin{align}
	\dot{\mathbf{x}} = g(\mathbf{x}, \theta, u)\\
	\mathbf{y} = h(\mathbf{x},\mu,u)
	\end{align}
\end{subequations}
When all states are not observable, Equation (\ref{eq:output}a) is usually augmented with a system that defines the relationship between the observable output variables ($\mathbf{y}$) and the state variables ($\mathbf{x}$) as in Equation (\ref{eq:output}b). The parameters used to establish this relationship (Equation \ref{eq:output}b), $\mu$, may or may not include system parameters $\theta$ defined in Equation (\ref{eq:output}a).

@@

\section{Methods:}	
\subsection{A method for practical identifiability of kinetic models of metabolism:}\label{sec:ident}
In this section, we show how practical identifiability of kinetic parameters in a dynamic model of metabolism can be established using the nonlinear algebraic kinetic rate law formulations that describe the relationship between the concentrations, fluxes and the kinetic parameters of the metabolic network model. A summary of the methodology in the form of a flow diagram is shown in Figure \ref{fig:ident-flowchart}.
In a kinetic model, the value of every flux $v_i$ is expressed using one of the many available enzyme kinetic formulations. Without loss of generality, all of these kinetic formulations can be expressed as nonlinear algebraic equations (Equation \ref{eq:flux_gen}). The fluxes are expressed as a function of the metabolite concentrations $\mathbf{x}$ and the kinetic parameters $\theta$ (Figure \ref{fig:ident-flowchart}a). 
\begin{align}\label{eq:flux_gen}
v_i = f(\mathbf{x},\mathbf{\theta}, u)
\end{align}
Let $\theta \in \mathbb{R}^p$ for each flux $v_i$ in the network. For each experiment $j = {1, 2, ..., n}$, we assume that all metabolite concentrations $\mathbf{x}$ and reaction fluxes $\mathbf{v}$ are measurable. The pertinent information for each experiment is available as a vector of concentrations and fluxes, $\mathbf{x}_j$ and $\mathbf{v}_j$, respectively (Figure \ref{fig:ident-flowchart}b).

In order to establish the identifiability of kinetic parameters for each flux $v_i$, we describe a computer algebra-based method. The primary use of the computer algebra system is to obtain closed form expressions for each parameter in $\theta$ for each flux $v_i$ (Figure \ref{fig:ident-flowchart}c). This is done by solving a system of nonlinear algebraic equations in $\mathbb{R}^p$, shown in Equation (\ref{eq:nonlineq}).
\begin{align}\label{eq:nonlineq}
v_{i, k} = f_k(\mathbf{x}_k,\theta, u_k) && \forall k=\{1, 2, ..., p\}\subset\{1, 2, ..., n\}
\end{align}
Each equation in (\ref{eq:nonlineq}), indicated by the index $k$, corresponds to the kinetic rate law expression $f(\mathbf{x},\mathbf{\theta}, u)$ for $v_i$, described earlier in Equation (\ref{eq:flux_gen}), written for concentrations and fluxes obtained from experiment $k$. Solving the system in Equation (\ref{eq:nonlineq}) results in $\mathbb{R}^p$ nonlinear expressions for parameters in $\theta$, where $N(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ is the numerator of $g$, and $D(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ is the denominator of $g$ (Figure \ref{fig:ident-flowchart}c).
\begin{align}\label{eq:theta-eq}
\theta_k = g_k(\mathbf{v}_i, \mathbf{x}, \mathbf{u}) = \frac{N_k(\mathbf{v}_i, \mathbf{x}, \mathbf{u})}{D_k(\mathbf{v}_i, \mathbf{x}, \mathbf{u})}
\end{align}
The identifiability of parameter $\theta_k$ for flux $v_i$ can be established by determining the value of $D_k(\mathbf{v}_i, \mathbf{x}, \mathbf{u})$ (Figure \ref{fig:ident-flowchart}c): any parameter $\theta_k$ is said to practically identifiable (practically non-identifiable) if $D_k(\mathbf{v}_i, \mathbf{x}, \mathbf{u})\neq0$ ($D_k(\mathbf{v}_i, \mathbf{x}, \mathbf{u}) = 0$).

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=.8\textwidth,height=.8\textheight,keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figures/ident_flow_chart}}
	\caption{A flow diagram showing the methodology developed to establish practical identifiability of parameters in kinetic models of metabolism. a) The steps for the construction of a kinetic model of a metabolic network are shown. The choice of rate law formulations to describe metabolic fluxes influences the identification methodology. The identifiability of parameters for each flux can be establish independently. b) in silico data generation to test the identifiability of a given parameter is an important step that is also required for experimental design using the proposed methodology. c) The steps for identifiability analysis for a parameters of a single flux are shown. d) If any of the proposed in silico experiments results in non-identifiability of any parameter, then the number and nature of experiments to be performed to establish identifiability of the corresponding parameter can be determined by following the steps shown.}\label{fig:ident-flowchart}
\end{figure}

\subsection{Experimental design for parameter estimation in kinetic models of metabolism:}
The identifiability of each parameter based on each experiment indexed as $j = \{1, ..., n\}$ is established based on the methodology described previously in Section \ref{sec:ident} and demonstrated in Section \ref{sec:example}. Subsequently, for any flux $v_i$, if for any $p$ combinations of indices $j$, the experimental concentrations ($\mathbf{x}_j$) and fluxes ($\mathbf{v}_j$) do not satisfy the condition for identifiability for any parameter in $\theta\in\mathbb{R}^p$, i.e., $D_k(\mathbf{v}_i, \mathbf{x}, \mathbf{u}) = 0$ for any $k$, then at least one of the $p$ experiments needs to be changed to make a parameter $\theta_k$ identifiable. Consequently, the corresponding experiment cannot be used for parameter estimation and needs to be discarded from the set of all necessary experiments. Furthermore, another experiment from $j = \{1, ..., n\}$ needs to be selected such that parameter $\theta_k$ is identifiable. This process has to be repeated until all parameters in $\theta\in\mathbb{R}^p$ are identifiable for flux $v_i$. In doing so, we can arrive at a set of $p$ experiments that will always result in practically identifiable parameters for flux $v_i$. This analysis can be performed for each flux in a metabolic network independent of all the other fluxes. Hence, our method is theoretically scalable even to genome-scale models. We show the application of this design methodology for flux $v_2$ in the gluconeogenic model in Section.

Note that if none of the $n$ pre-selected experiments satisfy the identifiability condition, then we can design an $(n+1)^{th}$ experiment that can replace one of the experiments that causes practical non-identifiability using our methodology.

In the following sections we provide a previously published kinetic model of a small gluconeogenic network, followed by a demonstration of our methodology to establish practical identifiability and experimental design for one of the fluxes in this network.

\subsection{Kinetic model of gluconeogenesis in E. coli:}\label{sec:small-model}
The proposed model for acetate consumption through gluconeogenesis and its corresponding kinetic model is used as a case study to illustrate the utility of identifiability analysis for the design of experiments for estimating parameters in kinetic models of metabolism. The kinetic model is described below.
\begin{equation}\label{eq:ode1}
\frac{d}{dt}pep=v_1-v_2-v_4
\end{equation}
\begin{equation}\label{eq:ode2}
\frac{d}{dt}fdp=v_2-v_3
\end{equation}
\begin{equation}\label{eq:ode3}
\frac{d}{dt}E=v_{e,max}\left(\frac{1}{1+\left(\frac{fdp}{K_{e}^{fdp}}\right)^{n_e}}\right) - d E
\end{equation}
The kinetic expressions for fluxes $v_1$ through $v_4$ are given below. The consumption of acetate through $v_1$ and conversion of \textit{pep} through $v_2$ are expressed in Equations (\ref{eq:flux1}) and (\ref{eq:flux2}) respectively using Michaelis-Menten kinetics. The acetate flux through $v_1$ is also governed by the quantity of available enzyme E. 
\begin{equation}\label{eq:flux1}
v_1 = k_{1}^{cat}E\frac{acetate}{acetate+K_{1}^{acetate}}
\end{equation}	
\begin{equation}\label{eq:flux2}
v_2 = V_{2}^{max}\frac{pep}{pep+K_{2}^{pep}}
\end{equation}
\begin{equation}\label{eq:flux3}
v_3 = V_{3}^{max}\frac{\tilde{fdp}\left(1+\tilde{fdp}\right)^3}{\left(1+\tilde{fdp}\right)^4+L_3\left(1+\frac{pep}{K_{3}^{pep}}\right)^{-4}}
\end{equation}
The allosterically regulated flux $v_3$ for the consumption of \textit{fdp} is expressed in Equation (\ref{eq:flux3}) using the Monod-Wyman-Changeux (MWC) model for allosterically regulated enzymes, where $\tilde{fdp}$ refers to the ratio of \textit{fdp} with respect to its allosteric binding constant $K_{3}^{fdp}$. The added flux $v_4$ for the export of \textit{pep} is expressed as a linear equation dependent on $pep$ in Equation (\ref{eq:flux4}).
\begin{equation}\label{eq:flux4}
v_4 = k_{4}^{cat}.pep
\end{equation}

@@

\section{Results:}
First, in Section \ref{sec:example} we demonstrate how our methodology can be applied for one of the fluxes in the gluconeogenic model. In Section \ref{sec:closedform} that follows, we discuss the ability to apply our methods to different nonlinear kinetic rate law formulations within the context of the gluconeogenic model. Then, in Section we discuss the different experimental design strategies that were found through our identifiability analysis to enable kinetic parameter estimation for the each flux in the model using steady state data. Finally, in Section we expand the application of our method to a relatively large (39 reactions) kinetic model of the red blood cell metabolism.

\subsection{Identifiability of parameters in a kinetic model of gluconeogenesis:}\label{sec:example}
Here, we demonstrate the use of our computer algebra-based methodology to establish practical identifiability of parameters for flux $v_2$ in the small model of gluconeogenesis described in Section \ref{sec:small-model}. Flux $v_2$ has two parameters, $V_2^{max}$ and $K_2^{pep}$ that need to be estimated from experimental data. Here, we assume that at least two different sets of experimental data for the concentrations and fluxes are available. We label the concentrations as ${pep}^1$ and ${pep}^2$, and the fluxes as ${v_2}^1$ and ${v_2}^2$ respectively, from each experiment. Accordingly, the nonlinear algebraic equations shown in Equation (\ref{eq:nonlineq}) can be formulated for flux $v_2$ as follows:
\begin{subequations}\label{eq:nonlin-flux2}
	\begin{align}
	{v_2}^1 = V_{2}^{max}\frac{pep^1}{pep^1+K_{2}^{pep}}\\
	{v_2}^2 = V_{2}^{max}\frac{pep^2}{pep^2+K_{2}^{pep}}
	\end{align}
\end{subequations}
Solving this simultaneous system of equations, we get two nonlinear algebraic equations in the parameters $V_2^{max}$ and $K_2^{pep}$ based on the form shown earlier in Equation (\ref{eq:theta-eq}).
\begin{subequations}\label{eq:v2_par}
	\begin{align}		
	V_2^{max} = \frac{v_2^1v_2^2(pep^1-pep^2)}{v_2^2pep^1-v_2^1pep^2}\\
	K_2^{pep} = \frac{pep^1(v_2^1pep^2-v_2^2pep^2)}{v_2^2pep^1-v_2^1pep^2}
	\end{align}
\end{subequations}
In Equation (\ref{eq:v2_par}), the denominator of the right hand side expression is used to test the identifiability of parameters $V_2^{max}$ (Equation \ref{eq:v2_par}a) and $K_2^{pep}$ (Equation \ref{eq:v2_par}b).

\subsection{Getting closed-form expressions for each flux in Kotte model:}\label{sec:closedform}
The complexity of the equations in our specific scenario is determined by the complexity of enzyme-metabolite interaction models used describe fluxes in metabolic networks. Although computer algebra systems (CAS) are capable of handling complex symbolic calculations, sometimes, the complexity of getting closed form expressions for all kinetic parameters of certain rate law formulations is too much for the CAS to overcome. We encountered this scenario in the case of the gluconeogenic model for flux $v_3$ where the kinetics of the allosterically activated reaction are described by the MWC model for allosteric regulation. In order to overcome this computational difficulty, we used a convenience kinetic rate law formulation to describe the allosteric interaction in $v_3$. We give this formulation in Section.

We believe this problem of computational tractability to occur in other complex kinetic rate law formulations as well. Examples?

\subsection{Experiments that can establish identifiability for each flux in Kotte model:}	
As indicated previously in Section, we performed different experiments by perturbing the parameters for fluxes $v_1$, $v_2$ and $v_3$ in the model. We also perturbed the input substrate concentration to collect data on the concentrations and fluxes within the network. We describe the results of using this data to establish the identifiability of various parameters in the kinetic model of the small metabolic network.

\subsection{Combinations of experiments that will enable identification of all model parameters:}
\subsection{Expanding methodology to RBC model:}

@@

@@
